Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India

### 2025-07-06 point-by-point-response.qmd

> - The authors should comment on the sample size for a pilot study.  For example, why did they select seven clusters? They should also comment on the number of residents and patients required to allow for a meaningful analysis of the feasibility outcomes.

We have added the rationale for the number of clusters and the number of units per cluster to the methods section. We did not conduct a formal power calculation, but deemed that the anticipated sample sizes would be sufficient to evaluate the feasibility outcomes.

> - The reason for the one-month lead in is unclear.  It is unclear how this data will be analyzed.

We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.

> - Using sealed envelopes for randomization increases the risk of selection bias, tampering, and accidental loss, potentially compromising allocation concealment. Authors should have used the computer-generated randomization system at time of randomization for each site.

We recognise that the use of sealed envelopes for randomisation increases the risk of bias and errors, and we have added this as a limitation of this pilot study.

> - For blinding, the authors should state that it was not possible to blind those involved.  It is unclear if those analyzing the data were blinded, as this is possible.

We have added that we did not blind the data analysts in this pilot study, but recommend doing so in the full scale trial.

> - The follow-up period for each participant is unclear. The methods state that data were collected for four months.  The methods also mention a 3-day mortality outcome.

The data collection period in each hospital was four months, but patients were followed up for one month. We have clarified this in the methods section.

> - Success criteria and threshold details for the feasibility outcomes should be clearly outlined in the methods.

We did not specify success criteria for the feasibility outcomes, and we have added this to the methods section.

> - The outcomes for the definitive trial (clinical outcomes) need to be clearly listed and described.

We now specify that we collected data on potential outcomes for the full scale trial, including mortality within 30 days, in-hospital mortality, complications and health related quality of life.

> - The statistical analysis section is unclear.  It should list the descriptive analyses plans and then the analyses used to assess the feasibility objectives. Feasibility pilot studies are not powered to detect meaningful differences between groups. Rather their goal is to determine whether a definitive trial is practical by assessing feasibility metrics. The clinical outcomes (the outcomes for the definitive trial) could be reported descriptively, but the use of comparative statistics (differences in final values, differences from baseline, confidence intervals, and subgroup analyses) to assess these outcomes is not appropriate during the pilot phase and potentially misleading, especially without a formal sample size calculation being conducted.

Thank you for pointing this out. We acknowledge that the estimation of potential effect sizes is not appropriate for this pilot study, and we have revised the aim of the study to reflect this and report this change as a deviation from the original protocol. We now report that we used descriptive statistics to analyse the feasibility outcomes and report clinical outcomes descriptively.

> - The statistical analyses section also describes inconsistent times for outcome comparisons. The intervention arms measure change from baseline over one month pre-training vs. three months post-training, while the control arm measures change over a four-month period. For the definitive trial, the timing of assessment should be consistent across all arms to avoid biased comparisons.

We agree.

> Results
> - Add a description of each cluster.

We have added a description of each cluster as Supplementary Table S1.

> - Comment on why the number of residents and patients varied across the treatment groups.

We have clarified that the number of residents trained in each intervention cluster depended on the size of the units, and that the number of patients included per cluster varied between clusters depending on the volume of patients at each hospital.

> - A few key prognostic factors differ across the treatment groups (age, ISS).

Thank you for pointing this out and we now highlight this in the results section.

> - Consider reporting overall enrollment rates (e.g. number of participants enrolled per month).  There is a paragraph in the discussion on low enrollment, but it is not listed in the results.

We have added a table to the results section to report the number of patients included per month.

> - The results of feasibility outcomes should be reported clearly in a table, with a column indicating whether the feasibility outcome was successfully achieved.

We have decided not to report the feasibility outcomes in a table, because some of them, for example the comparison of directly observed data and data extracted from medical records, were assessed qualitatively. We have however removed the description of the clinical outcomes from the outcomes section to avoid confusion.

> - Mortality should only be reported once in the results section.

Mortality is now reported together with the other patient characteristics in the results section. We have removed the table comparing clinical outcomes from the results section.

> - Report the resident outcomes and the participant outcomes separately.

We now report the resident outcomes and the participant outcomes separately.

> - List the protocol deviations in the results section, not in the discussion.

We now report the protocol deviations where relevant in the methods section, and include a list as supplementary material for completeness and easy reference.

> - It is unclear the importance and relevance of the before training after training data (Table S1).

We included the one month before training and three month after training data to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline. We have attempted to clarify this in the methods section.

> - The EQ-5D should be scored as opposed to showing the results for each question.

Currently, there is no Indian value set available for the EQ-5D-3L, and we have therefore not scored the data. We now explain this in the methods section.

> - Simplify the outcomes tables to focus on the key findings. Remove the yes/no and only list the yes.

We now only report the yes in the outcomes tables.

> - The absolute difference versus relative difference does not add to the analyses and could be removed.

We have removed the absolute difference versus relative difference from the main results section, but kept it in the supplementary materials as it was part of the original protocol.

> - As this a pilot study, outcome data should be reported descriptively in a clearer manner.

We have attempted to clarify the reporting of the outcome data in the results section.

> - There are too many outcomes.

We agree and suggest that the full scale trial should focus on the most relevant outcomes.

> Discussion
> - Remove all results from the discussion and place them in the results section, then discuss the results in the discussion section.

We now only discuss results already reported in the results section in the discussion.

> - Discuss the implication of adding the seventh site in more detail.

We have clarified that the seventh site was included for pragmatic reasons.

> - Discuss if there are methods to improve participant follow-up for the definitive trial.

We now highlight the challenges of following up patients after discharge in the discussion and suggest that the full scale trial should focus on in-hospital mortality as the primary outcome.

> - Discuss if there are methods to reduce missing data for the definitive phase.

We have added a suggestion to the discussion to improve the completeness of the data extracted from the medical records.

> - Comment on the ethics of the full trial. Is it ethical to have sites continue with usual care after seeing the pilot study results?

> - Consider a stepped wedge design where all clusters start with standard of care and then transition into an intervention.
> - Use the CONSORT statement extension for pilot and feasibility trials (https://www.equator-network.org/reporting-guidelines/consort-2010-statement-extension-to-randomised-pilot-and-feasibility-trials/) and the CONSORT statement extension for cluster randomised trials (https://www.equator-network.org/reporting-guidelines/consort-cluster/).

### 2025-07-06 manuscript.Rmd
Continue working with reviewer 3's comments

We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022]. There were a standard care arm and two intervention arms, ATLS^®^ and PTC training. We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm). In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [ADD CITATION].

@@

- Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all enrolled patients.
- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.
- Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
- Differences in distributions between directly observed data and data extracted from medical records. This applied to all variables that could be reasonably expected to be present in the medical records, but we only extracted this data from a convenience sample of patients to reduce the workload of the project officers collecting data.

We did not specify success criteria for these outcomes.

@@

Arriving patients were screened for eligibility and consented, if conscious. Unconscious patients were consented by the patient’s representative. This proxy consent was reaffirmed by the patient, on regaining consciousness. Patients who did not regain consciousness were included based on their representative's consent. We followed up patients at 24 hours after arrival at the emergency department, and at 30 days after arrival at the emergency department.

@@

Participating units were screened for eligibility once hospitals confirmed their participation. All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms. The protocol stated that residents would be approached for consent before the hospitals were randomised, but this proved not to be feasible, and we therefore asked residents for consent after the hospitals were randomised but before training. The training was conducted approximately one month after the study started in that hospital. Initially, we planned to use simple random sampling to select the units to be trained in each intervention hospital, but for pragmatic reasons we decided to leave the decision on which units to train to the site principal investigator.

@@

We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training. We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods. <!--The time period was dictated by budget and time constraints.-->

@@

We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. We did not blind investigators, residents or patients to the intervention.

@@

The research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained. All injuries were coded according to the International Classification of Diseases version 10 (ICD-10). Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022]. We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and cause of death. A list of variables is available in Supplementary Table S2.<!--Supplemental Material 2 contains a full variable list, with definitions.-->

@@

We enrolled `r n.patients` trauma patients from `r number_to_text(n.centres)` participating centres between `r start.date` and `r end.date`. The standard care arm enrolled `r n.control` patients, the ATLS^®^ arm enrolled `r n.atls` patients, and the PTC arm enrolled `r n.ptc` patients. We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS^®^, and  `r number_to_text(n.ptc.residents)` in PTC.

The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics). Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`. A total of `r m30d` patients died within 30 days after arrival to the emergency department, and `r inhosp` patients died in hospital.

@@

The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%. Among patients, the lost to follow up rate was `r rate.lost.to.follow.up`%. Among residents, the pass rate was `r pass.rate`%. The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2. The variables with the maximum amount of missing data were in the cost of treatment, reported in Supplementary Tables S1-12.

The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted). Overall, the data were similarly distributed, but there were considerably more missing values in the data extracted from medical records compared to the directly observed data.

@@

We show that it is feasible to conduct and collect data for a cluster randomized controlled trial comparing ATLS^®^ with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables, especially the cost of treatment (reported in Supplementary Table S2-13) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.

@@

All-cause 30-day mortality was missing in `r rate.lost.to.follow.up`% of patients. This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge. Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to previous trials. The difference in the number of patient deaths between the arms may have resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.

@@

There are some important limitations of this pilot study and therefore lessons to be learned and factored into the design of the full-scale trial. First, the volumes of patients included at some of the participating hospitals were lower than expected. A careful assessment of patient volumes as part of the screening process will be needed for the full-scale trial. Second, data on complications and cause of death were almost universally missing. To collect data on these variables will require other methods, as these were not explicitly described in the medical records and autopsy reports are not readily available. Third, we did not collect in-depth data on the standard care at each hospital. This data should be collected as part of the screening process for the full-scale trial.

>>>>

We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022]. There were a standard care arm and two intervention arms, ATLS^®^ and PTC training. We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm). We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline. In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [ADD CITATION].

@@

- Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all included patients.
- Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
- Differences in distributions between directly observed data and data extracted from medical records. This applied to all variables that could be reasonably expected to be present in the medical records, but we only extracted this data from a convenience sample of patients to reduce the workload of the project officers collecting data.
- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.

@@

Arriving patients were screened for eligibility and consented, if conscious. Unconscious patients were consented by the patient’s representative. This proxy consent was reaffirmed by the patient, on regaining consciousness. Patients who did not regain consciousness were included based on their representative's consent. We followed up patients at 24 hours after arrival at the emergency department, and at 30 days after arrival at the emergency department. The follow up period for each patient was therefore one month.

@@

Participating units were screened for eligibility once hospitals confirmed their participation. All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms. The protocol stated that residents would be approached for consent before the hospitals were randomised, but this proved not to be feasible, and we therefore asked residents for consent after the hospitals were randomised but before training. The training was conducted approximately one month after the study started in that hospital. Initially, we planned to use simple random sampling to select the units to be trained in each intervention hospital, but for pragmatic reasons we decided to leave the decision on which units to train to the site principal investigator. The number of residents trained in each intervention cluster varied because the size of the units varied.

@@

We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training. We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods. We anticipated that the numbers of patients included per cluster were anticipated would vary between clusters depending on the volume of patients at each hospital.<!--The time period was dictated by budget and time constraints.-->

@@

We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. It was not possible to blind investigators, residents or patients to the intervention. We did not blind the data analysts.

@@

The research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained. All injuries were coded according to the International Classification of Diseases version 10 (ICD-10). Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022]. We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and health related quality of life (using the EQ-5D-3L). We did not calculate an EQ-5D-3L index score, because there is currently no Indian value set available for the EQ-5D-3L [@jyani_development_2022]. We also attempted to collect data on cause of death. A list of variables is available in Supplementary Table S2.<!--Supplemental Material 2 contains a full variable list, with definitions.-->

@@

We included `r n.patients` trauma patients from `r number_to_text(n.centres)` clusters between `r start.date` and `r end.date`. The period of data collection and the number of patients included per month per cluster is shown in Table \@ref(tab:patients-per-month). The standard care arm included `r n.control` patients, the ATLS^®^ arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients. We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS^®^, and  `r number_to_text(n.ptc.residents)` in PTC.

```{r patients-per-month, echo = FALSE, results = "asis", tab.cap = "Number of patients included per month per cluster."}
patients.per.month.table |>
  modify_caption("Number of patients included per month per cluster.")
```

The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics). Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`. These prognostic factors differed between the trial arms. A total of `r m30d` patients died within 30 days after arrival to the emergency department, and `r inhosp` patients died in hospital.

@@

The percentage of patients who consented to follow up was `r p.consent`% and the lost to follow up rate was `r rate.lost.to.follow.up`%. The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2. The variables with the maximum amount of missing data were in the cost of treatment, reported in Supplementary Tables S1-12.

The differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted). Overall, the data were similarly distributed, but there were considerably more missing values in the data extracted from medical records compared to the directly observed data.

The percentage of residents who consented to training was `r recruitment.rate.residents`% and the pass rate was `r pass.rate`%.

@@

We show that it is feasible to conduct and collect data for a cluster randomized controlled trial comparing ATLS^®^ with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables, especially the cost of treatment (reported in Supplementary Table S2-13) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete. To improve completeness of data extracted from the medical records, a full scale trial should limit the number of variables extracted from the medical records and the importance of having these registered should be emphasised to the participating hospitals.

@@

All-cause 30-day mortality was missing in `r rate.lost.to.follow.up`% of patients. This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge. Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to previous trials. Following up patients after discharge is notoriously challenging in the setting of this study, and the full scale trial might need to focus on in-hospital mortality as the primary outcome. The difference in the number of patient deaths between the arms may have resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.

@@

There are some important limitations of this pilot study and therefore lessons to be learned and factored into the design of the full-scale trial. First, the volumes of patients included at some of the participating hospitals were lower than expected. A careful assessment of patient volumes as part of the screening process will be needed for the full-scale trial. Second, data on complications and cause of death were almost universally missing. To collect data on these variables will require other methods, as these were not explicitly described in the medical records and autopsy reports are not readily available. Third, we did not collect in-depth data on the standard care at each hospital. This data should be collected as part of the screening process for the full-scale trial. Fourth, we used sealed envelopes for randomisation, which increases the risk of bias and errors. The full scale trial should use a computer-generated randomisation system. Fifth, we did not blind the data analysts, but recommend doing so in the full scale trial. Finally, we assessed a very large number of potential outcomes, and the full scale trial should focus on the most relevant outcomes.
